### 2022-12-14 introduction/introduction.tex
Added paragraph about early disorder examples and tweaked wording in a few other places

Some examples will illustrate the central role of proteins more clearly. Blood is a part of the circulatory system which is responsible for transporting nutrients and waste. Though blood is a complex mixture, containing a cocktail of cells, proteins, sugars, gases, and ions dissolved in a medium of water, its primary cellular component is red blood cells. These cells, which give blood its red color, ferry oxygen from lungs throughout the body. While water can dissolve some oxygen, the body requires more oxygen more quickly than is available in the aqueous component of blood alone. Thus, red blood are packed with a special protein called hemoglobin, which binds\footnote{In biology, ``binding'' is a slippery word whose exact meaning can varying greatly depending on the context. However, it generally means that a molecule physically interacts with another molecular for an extended period.} oxygen. Each red blood cell contains as many as 270 million molecules of hemoglobin, each of which can carry up to four oxygen molecules. To put this value in perspective, if each molecule of hemoglobin were the size of a tennis ball, the hemoglobin molecules in a single red blood cell would cover a tennis court in over 10 packed layers\footnote{A standard tennis ball is 1.35 in or 0.1125 ft in radius, so the volume of 270 million tennis balls of this size packed as efficiently as possible is given by $2.70 \times 10^6 \times \frac{4}{3} \pi (0.1125 \text{ ft})^3 \times \frac{3\sqrt{2}}{\pi} = 2.17 \times 10^4 \text{ ft}^3$. A singles tennis court is 78 ft by 27 ft or 2106 ft\textsuperscript{2}, so dividing the first by the second yields slightly over 10 layers.}. Because red blood cells are so dense with hemoglobin, any defect in hemoglobin can dramatically impact the structure of the red blood cells themselves. A well-studied example is sickle cell disease where an error in the body's hemoglobin molecules deforms red blood cells into a characteristic sickle shape. This prevents them from easily flowing through blood vessels, resulting in pain and oxygen deprivation.

@@

Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group. However, proteins are only synthesized from the 20 ``canonical'' amino acids. Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.}. Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary receptor and donor sites for chemical bonds. Thus, in a protein the amino acids are bonded in a linear chain like beads on a string. However, once synthesized, proteins are not tidy rod-shaped molecules. Instead, the chain loops and weaves between itself creating a characteristic three-dimensional structure in a process called folding. These structures, which are highly stable and characteristic of each protein, are a result of the interactions between the amino acids in the chain and the surrounding medium, which is typically water. Because each amino acid has unique geometric and chemical properties that influence the energetics of these interactions, a protein's three-dimensional structure is encoded by the sequence of amino acids that compose it. A protein's function is in turn a direct result of its structure. For example, the structure of hemoglobin precisely positions its amino acids and a helper molecule called a heme group to create a pocket that can stably but reversibly bind oxygen. This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination. However, people affected by sickle cell disease have a mutation in the sequence of their hemoglobin proteins which causes it to malfunction. Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for a valine. This creates a sticky patch on the surface of hemoglobin, and under low-oxygen conditions normal hemoglobin changes shape to expose a sticky patch on its surface as well. The two patches are complementary, which allows hemoglobin proteins to clump together into long, fibrous strands. These strands distort the shape of red cells, giving them their characteristic sickle shape.

Because sequence dictates structure and structure dictates function, understanding these relationships in proteins is essential for understanding biological systems as a whole. While biologists are generally interested in function since it is the most directly associated with the biological role\footnote{Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function. For example, the strict function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin. However, the two have different roles. Hemoglobin is found in red blood cells where it primarily acts a carrier. In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.} of a protein, function and biological role are not always easily identified or measured. Thus, determining a protein's structure is frequently the first step of detailed study of a protein's function. Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods still remain the gold standard. While many methods can reveal information about the structure of a protein, the most powerful techniques, X-ray crystallography, NMR spectroscopy, and cryogenic electron microscopy (cryo-EM), can map the spatial coordinates of every atom in a protein. However, this resolution requires extremely pure samples of the protein of interest. Since proteins are only produced by living systems, preparations begin with a complex mixture consisting of cells or tissue, and the protein of interest is isolated with series of extraction and purification steps. Some proteins are only produced in small amounts or degrade easily, so each step may require substantial optimization to achieve a sufficient yield. When structural techniques were first developed in the late 1950s, they were so time-consuming that a graduate student could dedicate an entire PhD to solving a single protein structure. Many developments have substantially accelerated the process, but it remains a labor-intensive technique which may require several months of effort. However, the result is a powerful map that scientists use to suggest hypotheses and interpret data.

The success of structural methods through the latter half of the 20th century established the view that while proteins were not completely rigid and could adopt a variety of related structures or ``conformations,'' most functional proteins had a single dominant structure. While some proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, these were viewed as exceptions. However, in the late 1990s the first genomes were sequenced, which allowed scientists to analyse the complete set of protein sequences encoded by an organism's DNA for the first time.

When did scientists first notice the prevalence of IDRs?
What tools did they use to detect it?
In what organisms?

>>>>

Some examples will illustrate the central role of proteins more clearly. Blood is a part of the circulatory system which is responsible for transporting nutrients and waste. Though blood is a complex mixture, containing a cocktail of cells, proteins, sugars, gases, and ions dissolved in a medium of water, its primary cellular component is red blood cells. These cells, which give blood its red color, ferry oxygen from lungs throughout the body. While water can dissolve some oxygen, the body requires more oxygen more quickly than is available in the aqueous component of blood alone. Thus, red blood are packed with a special protein called hemoglobin, which binds\footnote{In biology, ``binding'' is a slippery word whose exact meaning can varying greatly depending on the context. However, it generally means that a molecule physically interacts with another molecular for an extended period.} oxygen. Each red blood cell contains as many as 270 million molecules of hemoglobin, each of which can carry up to four oxygen molecules~\cite{Pierig2008}. To put this value in perspective, if each molecule of hemoglobin were the size of a tennis ball, the hemoglobin molecules in a single red blood cell would cover a tennis court in over 10 packed layers\footnote{A standard tennis ball is 1.35 in or 0.1125 ft in radius, so the volume of 270 million tennis balls of this size packed as efficiently as possible is given by $2.70 \times 10^6 \times \frac{4}{3} \pi (0.1125 \text{ ft})^3 \times \frac{3\sqrt{2}}{\pi} = 2.17 \times 10^4 \text{ ft}^3$. A singles tennis court is 78 ft by 27 ft or 2106 ft\textsuperscript{2}, so dividing the first by the second yields slightly over 10 layers.}. Because red blood cells are so dense with hemoglobin, any defect in hemoglobin can dramatically impact the structure of the red blood cells themselves. A well-studied example is sickle cell disease where an error in the body's hemoglobin molecules deforms red blood cells into a characteristic sickle shape. This prevents them from easily flowing through blood vessels, resulting in pain and oxygen deprivation.

@@

Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids.\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group. However, proteins are only synthesized from the 20 ``canonical'' amino acids. Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.} Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary receptor and donor sites for chemical bonds. Thus, in a protein the amino acids are bonded in a linear chain like beads on a string. However, once synthesized, proteins are not tidy rod-shaped molecules. Instead, the chain loops and weaves between itself creating a three-dimensional structure in a process called folding. These structures, which are highly stable and characteristic of each protein, are a result of the interactions between the amino acids in the chain and the surrounding medium, which is typically water. Because each amino acid has unique geometric and chemical properties that influence the energetics of these interactions, a protein's three-dimensional structure is encoded by the sequence of amino acids that compose it. A protein's function is in turn a direct result of its structure. For example, the structure of hemoglobin precisely positions its amino acids and a helper molecule called a heme group to create a pocket that can stably but reversibly bind oxygen. This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination. However, people affected by sickle cell disease have a mutation in the sequence of their hemoglobin proteins which causes it to malfunction. Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for a valine. This creates a sticky patch on the surface of hemoglobin, and under low-oxygen conditions normal hemoglobin changes shape to expose a sticky patch on its surface as well. The two patches are complementary, which allows hemoglobin proteins to clump together into long, fibrous strands. These strands distort the shape of red cells, giving them their characteristic sickle shape.

Clearly, understanding the relationship between the sequence, structure, and function of proteins is essential for unraveling more complex biological phenomena. While biologists are generally most interested in the function of a protein since it is the most directly associated with its biological role\footnote{Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function. For example, the strict function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin. However, the two have different roles. Hemoglobin is found in red blood cells where it acts as a carrier during transport. In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.}, function and biological role are not always easily identified or measured. Thus, determining a protein's structure is frequently the first step of more detailed studies of its function. Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods still remain the gold standard. While many methods can reveal information about the structure of a protein, the most powerful techniques, X-ray crystallography, NMR spectroscopy, and cryogenic electron microscopy (cryo-EM), can map the spatial coordinates of every atom in a protein. However, this resolution requires extremely pure samples of the protein of interest. Since proteins are only produced by living systems, preparations begin with a complex mixture consisting of cells or tissue, and the protein of interest is isolated with series of extraction and purification steps. Some proteins are only produced in small amounts or degrade easily, so each step may require substantial optimization to achieve a sufficient yield. When structural techniques were first developed in the late 1950s, they were so time-consuming that a graduate student could dedicate an entire PhD to solving a single protein structure. Many developments have substantially accelerated the process, but it remains a labor-intensive technique which may require several months of effort. However, the result is a powerful map that scientists use to suggest hypotheses and interpret data.

The success of structural methods at elucidating the molecular details of protein function cemented the view that function largely depends on the presence of a fixed structure. While scientists understood proteins were not completely rigid and could adopt a variety of related structures or ``conformations,'' many believed that functional proteins largely had a single dominant structure. Despite its strength, exceptions to the structure-function paradigm were known. For example, elastin is a protein secreted by cells which allows tissues like skin or blood vessels to repeatedly expand and contract. It imparts this elasticity by forming networks of disordered chains that act like molecular springs. When a tissue experiences a force, the chains stretch to accommodate it. When the force is removed, the chains return to their random orientations, which reduces their end-to-end length and forces the tissue to contract~\cite{Vrhovski1998, Alberts2014}. However, as a result of this unique role in providing tissue structure, elastin's disorder was viewed as a specific adaptation rather than a general mode of protein function. In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, but because these regions were often short loops between structured regions, they were seen as ``filler'' which facilitated the structure of the functional portions of the protein. By the late 1990s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Wright1999}.
